\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-15 submission/manuscript.Rmd
discussion

<!-- Context -->
Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2019phenomenal; @makowski2018cognitive]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus, and its credibility, authority and expertise, to help facilitate their evaluation [petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) can render this task difficult [*REF*]. In the absence of contextual information, what drives our beliefs of reality?

@@

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of faces evaluation, such as trustworthiness or attractiveness (**REF**). Interestingly, several studies report that artificially created faces were judged to be less attractive (**QUESTION: was their nature known to the participant?**) as compared to real faces [@tsikandilakis2019beauty; @diel2022familiarity], and @liefooghe2022faces further reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that beliefs of reality have an effect on attractiveness ratings of faces, the opposite question, whether attractiveness could drive beliefs of reality, has received no attention to date.

<!-- Goals and hypotheses -->
This study aims primarily at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. Moreover, we expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the role of dispositional traits, such as personality and attitude towards AI, characteristics on inter-individual simulation monitoring tendencies.

@@

Although expected, the first striking result is that despite all items being real photos from the same database, all people easily believe (to high degrees of confidence) that a significant proportion of them was fake. This is a testimony to the existing expectations regarding CGI technology, as much as to the volatility of our sense of reality. In fact, stimuli-related and participant-related characteristics accounted for about only less than 20% of the variance in the beliefs, suggesting that a large part of it is related to other subjective processes.

Though attractiveness does not seem to be the primary drive underlying simulation monitoring of face images, it does nonetheless display an association.

>>>>

<!-- Context -->
Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2019phenomenal; @makowski2018cognitive]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus, and its credibility, authority and expertise, to help facilitate their evaluation [@petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) can render this task difficult [@berghel2018weaponizing; @chen2015news]. In the absence of contextual information, what drives our beliefs of reality?

@@

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of faces evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, some studies report that when the nature of the faces was ambiguous, artificially created faces that were previously rated as more attractive were judged by subjects to be less real [@tucciarelli2020]. However, as the attractiveness ratings were given by independent raters instead of the participants, the direction of the relationship between perceived realness and attractiveness cannot be concluded. To this end, @liefooghe2022faces reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that beliefs of reality have an effect on attractiveness ratings of faces, the opposite question, whether attractiveness could drive beliefs of reality, has received little attention to date.

<!-- Goals and hypotheses -->
This study aims primarily at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. Moreover, we expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the role of dispositional traits, such as personality and attitude towards AI, as well as characteristics of inter-individual simulation monitoring tendencies.

@@

Notably, despite all the facial images being of real individuals retrieved from the same database, our results found every participant easily believed (to high degrees of confidence) that a significant proportion of them were fake. While this result is unsurprising, it is nevertheless a testimony to the existing expectations regarding CGI technology, as well as to the volatility of our sense of reality. In fact, stimuli-related and participant-related characteristics accounted for less than 20% of the variance in beliefs, suggesting that a large part of it is associated with other subjective processes.

### Attractiveness and Beauty (males -positive linear; females -quadratic(u-shape); judgement confidence positive quadratic relationship for females only in the real condition)

Although attractiveness did not seem to be the primary drive underlying simulation monitoring of facial images in the present study, it does nonetheless display an association. Furthermore, the positive quadratic relationship found between beliefs of realness and attractiveness for female participants is aligned with our hypothesis, that faces which are highly salient (i.e., perceived to be very attractive or very unattractive) are judged to be more real. Alternatively, the positive linear relationship between beauty and attractiveness, and beliefs of realness observed for males could have an evolutionary basis. In particular, males are widely postulated to place more importance toward facial attractiveness as a sign of reproductive potential, relative to females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender]. As such, males could have perceived highly attractive and beautiful faces to be more real since such attractiveness act as salient cues of biological quality [@fink2006facial]. Therefore, our study suggests a more complex relationship between stimulus attractiveness, demographic variables and perceptions of reality that has not been examined in the literature.

Moreover, beyond its impacts on individuals' beliefs about realness, attractiveness was found to significantly affect the confidence in judgements for females.

### trustworthiness

In contrast to the findings for attractiveness, trustworthiness only positively predicted beliefs in realness for females. While we had expected a positive relationship between the two variables across both genders, this observation is also consistent with the evolutionary perspective, which suggests that females, relative to males, are more averse to risk-taking because it may compromise their reproductive potential [@van2020sex]. As such, given that low facial trustworthiness may signal greater risk, perceived trustworthiness may confer greater importance in predicting the perception of reality for females than males [@hou2019survival]. Interestingly, while several studies have found both possessing prior knowledge that faces were artificial and the perceived realness of virtual faces to be predictive of trustworthiness ratings (i.e., faces judged as real were rated as more trustworthy) [@liefooghe2022faces; @hoogers2021effect; @balas2017], the same cannot easily be said of the opposite direction; that is, trustworthiness does not seem to be a primary driver underlying the processes of simulation monitoring.

### familiarity (confidence judgement; positive linear -real; quadratic -fake)
Contrary to our hypothesis, familiarity was not found to have any significant effects on simulation monitoring decisions.

### inter-indiv correlates (HH- negative with confidence in real; positive with acclaim-seeking and grandiose; Acc-seeking - positive with confidence in fake)

### AI attitudes (enthusiam - positive with confidence in both real and fake judgements)

### Re-exposure Delay

Although the order of presentation of the facial images was randomized to reduce effects of adaptation, re-exposure delay was found to have a significant negative effect on simulation monitoring decisions, with shorter delays being associated with faces being rated as more real. Interestingly, while it could be posited that shorter delays led to the faces appearing more familiar and thereby increased people's belief in its realness, this seems unlikely considering perceived familiarity did not significantly affect simulation monitoring decisions made, even after controlling for attractiveness and trustworthiness. Alternatively, shorter re-exposure delays could have led to the faces being better remembered, thus triggering autobiographical memory processes [@gobbini2013prioritized] and evoking a sense of personal relevance during the repeated display. Indeed, fictional stimuli that were associated with more personal memories have been shown to up-regulate emotions [@sperduti2016paradox; @makowski2017being], thus biasing the realness of the given stimulus
